---
title: "DS 202 Final Project Proposal"
author: "Jesse Dolan, Bela Banegas, Jennifer Godbersen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Research Project Title: Data Science Salaries

This data set includes information about data science salaries. Variables include: year, experience level, employment type (full-time, part-time, contract, freelance), job title, salary, salary currency, salary in usd, employee residence, remote ratio (amount of work done remotely), and company location.

```{r}
file_path <- "salaries.csv"

df <- read.csv(file_path)

head(df)

```

Data cleaning steps: There do not appear to be any empty rows or NA values.

```{r}
empty_rows <- df[!complete.cases(df), ]
if (any(is.na(df))) {
  print("There are NA values in the dataframe.")
} else {
  print("There are no NA values in the dataframe.")
}
```

Our current project idea is to identify the current salaries for people in our field. Particular variables of interest are experience level, job title, remote work ratio, and company title.
